Entropy Rain: Dilution and Compression of Thermals in Stratified Domains

### 2019-07-27 tex/stratified_thermals.tex
fixes FC nondimensionalization in text...need to make sure that typo doesn't make it through proofing.

\received{\today}
\revised{??}
\accepted{??}%\today}
\submitjournal{ApJ}

@@

\begin{abstract}
Large-scale convective flows called giant cells were once thought to transport the Sun's luminosity in the solar convection zone, but recent observations have called their existence into question.
In place of large-scale flows, some authors have suggested the solar luminosity may instead be transported by small droplets of rapidly falling, low entropy fluid.
This ``entropy rain'' could propagate as dense vortex rings, analogous to rising buoyant thermals in the Earth's atmosphere.
In this work, we develop an analytical theory describing the evolution of dense, negatively buoyant thermals.
We verify the theory with 2D \textbf{cylindrical} and 3D \textbf{cartesian} simulations of laminar, axisymmetric thermals in highly stratified atmospheres.
Our results show that dense thermals fall in two categories: a \emph{stalling} regime in which the droplets slow down and expand, and a \emph{falling} regime in which the droplets accelerate and shrink as they propagate downwards.
We estimate that solar downflows are in the falling regime and maintain their entropy perturbation against diffusion until they reach the base of the convection zone.
This suggests that entropy rain may be an effective nonlocal mechanism for transporting the solar luminosity.
\end{abstract}

@@

Thermals in the Boussinesq limit have been well studied in the laboratory for decades \citep[see e.g.][]{morton&all1956, scorer1957}, and more recently through Direct Numerical Simulation \citep[DNS,][]{lecoanet&jeevanjee2018}. 
These studies find that thermals expand radially and decelerate as they propagate.
Such a deceleration may cause the thermals to move so slowly they would diffuse away their entropy perturbation before reaching the bottom of the solar convection zone.
\textbf{\citet{brandenburg&hazlehurst2001} found that hot, buoyant thermals in stratified domains behave qualitatively similar to Boussinesq thermals.
However, we are not aware of past work which carefully examines the effects of stratification on \emph{negatively} buoyant thermals.
}

@@

Ignoring entropy variations, \citet{brandenburg2016} suggests the filling fraction $f$ of entropy rain should decrease like $f \propto \rho^{-1}$ for horizontal compression and $f \propto \rho^{-2/3}$ for spherical compression, \textbf{where $\rho$ is the density}.
On the other hand, the filling fraction of Boussinesq thermals \emph{increases} like $f \propto d^2$, where $d$ is the depth of the thermal.
These regimes are shown in Fig.~\ref{fig:overview}, and compared to the true propagation of a numerically simulated dense thermal which includes both entropy variations and density stratification.

@@

\subsection{Phenomenological description of thermal evolution}
\textbf{In} Fig.~\ref{fig:evolution_colormeshes}, \textbf{we show snapshots of 3D simulation data depicting} the descent of cold thermals released from rest in two domains which span a different number of density scale heights, $N_\rho$.
The left panel shows a weakly stratified domain with $N_\rho=0.5$, whereas the right panel shows a strongly stratified domain with $N_\rho=3$.
In both cases, the thermal is initialized with a spherical negative entropy perturbation whose diameter is 5\% of the domain depth.
This dense sphere spins up into an axisymmetric vortex ring, and the vertical cross section through this vortex ring shows two circular vorticity and entropy extrema.
In the $N_\rho = 0.5$ simulation, the thermal grows with depth, similar to thermals in the Boussinesq regime.
On the other hand, in the $N_\rho = 3$ simulation, the thermal's radius decreases marginally with depth.

@@

In the following sections, we will use the impulse of dense vortex rings to derive expressions for the evolution of their depth and radii with time.
\textbf{In this work we study vortex rings generated by discrete cold thermals, but ``plumes'' driven by time-stationary cooling produce similar vortex ring structures \citep[as in e.g.,][]{rast1998}.
The following description of vortex ring evolution should therefore be broadly applicable.}

\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{evolution_colormeshes.pdf}
    \caption{
	\textbf{Shown is the evolution of entropy perturbations, normalized by the thermal's buoyancy perturbation, from 3D simulations conducted in this work}.
	(left) A thermal in a weakly stratified domain with $N_\rho = 1/2$ density scale heights, and (right) a thermal in a strongly stratified domain with $N_\rho = 3$.
	While both start with precisely the same initial condition, the thermal in low stratification expands with depth \textbf{and slows down}, whereas the thermal in strong stratification compresses with depth \textbf{and accelerates}.
    \label{fig:evolution_colormeshes} }
\end{figure}

@@

The hydrodynamic impulse is defined as \citep{shivamoggi2010},
\begin{equation}
\bm{I} = \frac{1}{2}\int_{\mathcal{V}} \bm{x}\times(\grad\times(\rho\bm{u}))dV,
\end{equation}
where $\bm{x}$ is the position vector \textbf{and $\bm{u}$ is the fluid velocity}. 
The impulse is the time-integrated work which has acted on the fluid to result in the current fluid motion.
It is thus unaffected by internal forces (e.g., pressure or viscous).
Upon integration by parts, it is obvious that the impulse encompasses the momentum of the fluid within the volume $\mathcal{V}$.
However, one can also show that the surface terms correspond to the momentum outside the volume $\mathcal{V}$ \citep[e.g.][]{Akhmetov2009}.
A thermal with volume $\mathcal{V}$, density $\rho$, and translating with velocity $\bm{u} = w_{\text{th}} \hat{z}$ thus has an impulse
\begin{equation}\label{eqn:momentum}
I_z = (1+k) \rho \mathcal{V} w_{\text{th}},
\end{equation}
where $k$ encompasses the ``virtual mass effect'' from the environmental fluid moving together with the thermal \citep{tarshish&all2018}.

We now restrict our study to an ideal gas in the low Mach-number regime, in an adiabatic background.
\textbf{This is the appropriate regime of deep solar convection.}
Due to rapid pressure equilibration in low Mach-number flows, we can approximate 
\begin{equation*}
\frac{\rho_1}{\rho_0} \approx -\frac{S_1}{c_P},
\end{equation*}
where $S_1$ is the specific entropy perturbation and $c_P$ is the specific heat at constant pressure; thermodynamic variables are decomposed into background (subscript 0) and fluctuating (subscript 1) components.

The rate of change of the impulse is
\begin{equation*}
\frac{d\bm{I}}{d t} = \int_\mathcal{V} \rho_1\, \bm{g}\, dV,
\end{equation*}
because the surface terms completely cancel.
Assuming \textbf{a uniform, vertical gravity,} $\bm{g} = -g\hat{z}$, it is useful to define the buoyancy perturbation,
\begin{equation}
B \equiv \int_{\mathcal{V}} \rho_0\, S_1\, \frac{g}{c_P}\, dV.
\label{eqn:tot_buoyancy}
\end{equation}
such that
\begin{equation}
\frac{d I_z}{d t} = B.
\label{eqn:change_in_impulse}
\end{equation}

@@

\subsection{Model of thermal evolution}
\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{constants.pdf}
    \caption{
	\textbf{Shown is the} evolution of buoyancy perturbation (top), circulation (middle), and volume factor (bottom) \textbf{for each 2D simulation conducted in this work}.
	All \textbf{three of these quantities remain} nearly constant after an initial spin-up phase.
	With increasing stratification, we see marginally more detrainment \textbf{(loss of buoyant signature)} in the \textbf{top panel}.
    \label{fig:constants} }
\end{figure}

%After spinning up, we assume that the thermal has a \textbf{constant} negative buoyancy \textbf{perturbation} and \textbf{constant} circulation.
\textbf{The thermals studied here} began as initial spherical perturbations \textbf{and spin up into vortex rings}.
\textbf{The spun up vortex ring phase can be modeled as having evolved} from a ``virtual origin'' where the vortex ring had zero radius.
We model the thermal as having been located at its virtual origin at \textbf{a temporal offset} $t = -t_{\text{off}}$, where $t = 0$ is the time at which the true thermal was released from rest.

\textbf{In our simulations,} we find the thermal undergoes weak detrainment \textbf{(loss of buoyant signature to environmental fluid)}, so the negative buoyancy of the thermal decreases slightly in time (Fig.~\ref{fig:constants}, top panel).
We thus express the buoyancy perturbation as 
\begin{equation}
B \approx \eta B_{\text{th}},
\end{equation}
where $\eta$ is a constant of $\mathcal{O}$(1) which represents this detrainment \textbf{and $B_{\text{th}}$ is the thermal's characteristic buoyancy perturbation}. 
We then integrate Eqn.~\ref{eqn:change_in_impulse},
\begin{equation*}
I_z = \eta B_{\text{th}} (t + t_{\text{off}}).
\end{equation*}

Combining this with Eqn.~\ref{eqn:impulse_approx}, we retrieve our first main result,
\begin{equation}
r = \sqrt{\frac{\eta B_{\text{th}} (t + t_{\text{off}})}{\pi\rho_0\Gamma_{\text{th}}}},
\label{eqn:r_theory}
\end{equation}
\textbf{Here, $\Gamma_{\text{th}}$ is the characteristic circulation of the thermal. As} there are no net baroclinic torques, $\Gamma_{\text{th}}$ \textbf{remains nearly constant in our simulations} (Fig.~\ref{fig:constants}, middle panel).
In the Boussinesq limit where $\rho \rightarrow \text{constant}$, we retrieve the $r \propto \sqrt{t}$ scaling found in \LJ.
In the limit of strong stratification, we find $r \propto \rho^{-1/2}$, corresponding to purely horizontal compression, with $r^2 \propto \rho^{-1}$ \citep{brandenburg2016}.

To solve for the vertical evolution of the thermal, we can use Eqn.~\ref{eqn:momentum}.
We approximate the volume as $\mathcal{V} \approx m r^3$, where $m$ is a parameter which we take to be constant (which is a decent assumption \textbf{in our simulations}; see Fig.~\ref{fig:constants}, bottom panel).
\textbf{Here, $m$ accounts for volumetric constants (e.g., $4\pi/3$), the aspect ratio of the thermal, and the cubed ratio between the full radius of the spheroidal thermal and $r$.}
Defining the thermal velocity $w_{\text{th}} \equiv dz_{\text{th}}/dt$, we find
\begin{equation}
\frac{dz_{\text{th}}}{\rho(z_{\text{th}})^{1/2}} =
\left(\frac{(\pi\Gamma_{\text{th}})^{3/2}}{m(1 + k)(\eta B_{\text{th}})^{1/2}}\right)\frac{dt}{(t + t_{\text{off}})^{1/2}}.
\label{eqn:dz_theory}
\end{equation}
This can be easily integrated given an atmospheric stratification $\rho(z)$.

@@

\subsection{Polytropic atmosphere solution}
In this work, we study an ideal gas with an adiabatic index of $\gamma = 5/3$.
An adiabatic polytrope satisfies
\begin{gather}\label{eqn:T0}
T_0 = 1 + (z - L_z)\grad_{\text{ad}}, \\
\rho_0 = T_0^{\,n_{\text{ad}}},
\label{eqn:polytrope}
\end{gather}
where $n_{\text{ad}} = (\gamma-1)^{-1}$ \textbf{and $\grad_{\text{ad}}$ is the adiabatic temperature gradient}.
All thermodynamic quantities are nondimensionalized such that $\rho_0 = T_0 = 1$ at $z = L_z$, the top of the domain.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% EXPERIMENT SECTION
\section{Simulation setup} 
\label{sec:experiment}
To test our theory, we run a series of \textbf{thermal} simulations \textbf{using} the 3D fully compressible equations \textbf{in cartesian domains}.
\textbf{We additionally compute 2D azimuthally-symmetric} simulations \textbf{using} the anelastic equations \textbf{in cylindrical domains}.
We verify the 3D and 2D simulations produce the same results when run with the same parameters.
Because 2D simulations are less computationally expensive, we use them to cover a broader parameter regime.

@@

Following \LJ\,, we non-dimensionalize the equations \textbf{on} the initial diameter of the thermal and its freefall velocity.
These equations are then fully specified in terms of the Reynolds number and Prandtl number,
\begin{equation}
\text{Re} = \frac{ u_{\text{th}} L_{\text{th}}}{\nu}, \qquad
\text{Pr} = \frac{ u_{\text{th}} L_{\text{th}}}{\chi}, \qquad
u_{\text{th}}^2 = \frac{g L_{\text{th}} \Delta s}{c_P},
\end{equation}
where $u_{\text{th}}$ is the freefall velocity, $L_{\text{th}}$ is the thermal length scale, and
$\Delta s$ is the magnitude of the specific entropy signature of the thermal.

The background density and temperature are given by Eqs.~\ref{eqn:T0} \& \ref{eqn:polytrope}.
The adiabatic temperature gradient is $\nabla_{\text{ad}}=g(e^{N_\rho/n_{\text{ad}}}-1)/(L_z c_P)$, where $L_z=20$ is the height of the domain \textbf{and $N_\rho$ is the number of density scale heights spanned by the domain}.

We choose an atmospheric model in which the dynamic viscosity, $\mu = \rho_0 \nu$, and the thermal conductivity, $\kappa = \rho_0 \chi$, are both uniform \textbf{in space} and constant in time.
We make this choice because $\mu$ and $\kappa$ appear in the density-weighted momentum and entropy equations, and we find the density-weighted entropy and momentum to be key quantities in our thermal theory.
The diffusivities $\nu$ and $\chi$ therefore scale inversely with the density.
As the diffusivities scale with depth, Re is specified at the thermal's initial depth.
All simulations conducted in this work use \textbf{an initial} value of Re = 600 and $\mbox{\text{Pr} = 1}$.

\subsection{3D Fully Compressible Simulations}
In order to verify our 2D anelastic simulations, we also simulate thermals with the 3D Navier Stokes equations. 
We use the $(T, \ln\rho)$ formulation of the equations \citep{lecoanet&all2014, anders&brown2017},
\begin{gather}
\frac{\partial \ln\rho_1}{\partial t} + \epsilon^{-1}\left(\bm{u}\cdot\grad\ln\rho_0 + \DivU\right) = -\bm{u}\cdot\grad\ln\rho_1, \\
\begin{split}
\frac{\partial \bm{u}}{\partial t}  +\grad T_1 + &T_1\grad\ln\rho_0 + T_0\grad\ln\rho_1  =\\
- \epsilon &T_1\grad\ln\rho_1 + \frac{1}{\rho\text{Re}}\left[\grad^2\bm{u} + \frac{1}{3}\grad(\DivU)\right]
\end{split} \\
\begin{split}
&\frac{\partial T_1}{\partial t} + \epsilon^{-1}\left[\bm{u}\cdot\grad T_0 + (\gamma-1)T_0\DivU\right] = \\
&-\left[\bm{u}\cdot\grad T_1 + (\gamma-1)T_1\DivU\right] + \frac{1}{\rho c_V\text{Re}}\left[\frac{1}{\text{Pr}}\grad^2 T_1 + \sigma_{ij}\partial_{x_i}u_j\right].
\end{split}
\end{gather}
These equations are non-dimensionalized in the same way as the anelastic equations, and use the same background atmosphere.
The new parameter $\epsilon = u_{\text{th}}^2$ is the magnitude of entropy perturbations and sets the Mach number of the thermal flows analagously to the adiabatic excess in polytropic convection \citep{anders&brown2017}; we use $\epsilon = 10^{-4}$ in this work.

@@

For the fully compressible simulations, we must also specify the density perturbation $\rho_1$.
We pick perturbations that are in pressure equilibrium,
\begin{equation}
\ln\rho_1 = S_1/c_P, \qquad T_1 = T_0(e^{-\epsilon\ln\rho_1} - 1)/\epsilon.
\end{equation}
\textbf{In all cases, we do not add any symmetry breaking perturbations (e.g., noise).}

@@

The 2D simulation domain is \textbf{periodic in the z-direction with $z \in [-L_z/4, L_z]$ and the radial direction spans $r \in [0, L_r]$}.
The boundary conditions are $\partial_r S_1 = w = (\grad\times\bm{u})_\phi = 0$ at $r = L_r$, and the regularity of the equations automatically impose $u = \partial_r(w) = \partial_r(S_1) = 0$ at $r = 0$.
The 3D simulation domain is \textbf{periodic in the horizontal directions ($x, y \in [-L_r, L_r]$) and vertically spans $z \in [0, L_z]$}.
We impose impenetrable, stress free, fixed-temperature boundary conditions at the upper and lower boundaries.
In all of our simulations we specify $L_z = 20$ and $L_r = 5$.
We extend our 2D simulation domains to $z = -L_z/4$ because those simulations are vertically periodic.
This extension allows us to study the full transit of the thermal above $z = 0$ and terminate the simulation before it begins to wrap through the bottom of the periodic domain.

\textbf{
In our 2D simulations, we represent the radial direction with Chebyshev polynomials so that we can include geometric factors in our equations and capture the singularity at $r = 0$.
We then assume the z direction is periodic to make the calculations more efficient, and we end the simulations before the thermal can interact with the vertical periodic boundaries.
Remarkably, the results of these simulations vary only minimally from our 3D simulations, as we will show in the next section, indicating that the vertical periodicity and the different side boundaries (cylinder vs. cube) do not influcence the thermal properties.
}

All of the code used to perform the simulations in this work can be found online in the supplementary materials in a Zenodo repository \citep{supp_andersetall2019} \textbf{at \dataset[10.5281/zenodo.3311894]{https://doi.org/10.5281/zenodo.3311894}}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% RESULTS SECTION
\section{Model verification}
\label{sec:results}
\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{results_panels.pdf}
    \caption{
	(top) The thermal depth as a function of time for the 2D anelastic \textbf{(AN)} and 3D fully compressible \textbf{(FC)} simulations.
	(bottom) The corresponding thermal velocities as a function of depth.
	Theoretical predictions from section \ref{sec:theory} are plotted in thin solid lines for each case (see parameters in table \ref{table:parameters}).
    \label{fig:results_panels} }
\end{figure}

>>>>

\received{2019 June 6}
\revised{2019 July 25}
\accepted{2019 July 26}
\submitjournal{ApJ}

@@

\begin{abstract}
Large-scale convective flows called giant cells were once thought to transport the Sun's luminosity in the solar convection zone, but recent observations have called their existence into question.
In place of large-scale flows, some authors have suggested the solar luminosity may instead be transported by small droplets of rapidly falling, low entropy fluid.
This ``entropy rain'' could propagate as dense vortex rings, analogous to rising buoyant thermals in the Earth's atmosphere.
In this work, we develop an analytical theory describing the evolution of dense, negatively buoyant thermals.
We verify the theory with 2D cylindrical and 3D cartesian simulations of laminar, axisymmetric thermals in highly stratified atmospheres.
Our results show that dense thermals fall in two categories: a \emph{stalling} regime in which the droplets slow down and expand, and a \emph{falling} regime in which the droplets accelerate and shrink as they propagate downwards.
We estimate that solar downflows are in the falling regime and maintain their entropy perturbation against diffusion until they reach the base of the convection zone.
This suggests that entropy rain may be an effective nonlocal mechanism for transporting the solar luminosity.
\end{abstract}

@@

Thermals in the Boussinesq limit have been well studied in the laboratory for decades \citep[see e.g.][]{morton&all1956, scorer1957}, and more recently through Direct Numerical Simulation \citep[DNS,][]{lecoanet&jeevanjee2018}. 
These studies find that thermals expand radially and decelerate as they propagate.
Such a deceleration may cause the thermals to move so slowly they would diffuse away their entropy perturbation before reaching the bottom of the solar convection zone.
\citet{brandenburg&hazlehurst2001} found that hot, buoyant thermals in stratified domains behave qualitatively similar to Boussinesq thermals.
However, we are not aware of past work which carefully examines the effects of stratification on \emph{negatively} buoyant thermals.

@@

Ignoring entropy variations, \citet{brandenburg2016} suggests the filling fraction $f$ of entropy rain should decrease like $f \propto \rho^{-1}$ for horizontal compression and $f \propto \rho^{-2/3}$ for spherical compression, where $\rho$ is the density.
On the other hand, the filling fraction of Boussinesq thermals \emph{increases} like $f \propto d^2$, where $d$ is the depth of the thermal.
These regimes are shown in Fig.~\ref{fig:overview}, and compared to the true propagation of a numerically simulated dense thermal which includes both entropy variations and density stratification.

@@

\subsection{Phenomenological description of thermal evolution}
In Fig.~\ref{fig:evolution_colormeshes}, we show snapshots of 3D simulation data depicting the descent of cold thermals released from rest in two domains which span a different number of density scale heights, $N_\rho$.
The left panel shows a weakly stratified domain with $N_\rho=0.5$, whereas the right panel shows a strongly stratified domain with $N_\rho=3$.
In both cases, the thermal is initialized with a spherical negative entropy perturbation whose diameter is 5\% of the domain depth.
This dense sphere spins up into an axisymmetric vortex ring, and the vertical cross section through this vortex ring shows two circular vorticity and entropy extrema.
In the $N_\rho = 0.5$ simulation, the thermal grows with depth, similar to thermals in the Boussinesq regime.
On the other hand, in the $N_\rho = 3$ simulation, the thermal's radius decreases marginally with depth.

@@

In the following sections, we will use the impulse of dense vortex rings to derive expressions for the evolution of their depth and radii with time.
In this work we study vortex rings generated by discrete cold thermals, but ``plumes'' driven by time-stationary cooling produce similar vortex ring structures \citep[as in e.g.,][]{rast1998}.
The following description of vortex ring evolution should therefore be broadly applicable.

\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{evolution_colormeshes.pdf}
    \caption{
	Shown is the evolution of entropy perturbations, normalized by the thermal's buoyancy perturbation, from 3D simulations conducted in this work.
	(left) A thermal in a weakly stratified domain with $N_\rho = 1/2$ density scale heights, and (right) a thermal in a strongly stratified domain with $N_\rho = 3$.
	While both start with precisely the same initial condition, the thermal in low stratification expands with depth and slows down, whereas the thermal in strong stratification compresses with depth and accelerates.
    \label{fig:evolution_colormeshes} }
\end{figure}

@@

The hydrodynamic impulse is defined as \citep{shivamoggi2010},
\begin{equation}
\bm{I} = \frac{1}{2}\int_{\mathcal{V}} \bm{x}\times(\grad\times(\rho\bm{u}))dV,
\end{equation}
where $\bm{x}$ is the position vector and $\bm{u}$ is the fluid velocity. 
The impulse is the time-integrated work which has acted on the fluid to result in the current fluid motion.
It is thus unaffected by internal forces (e.g., pressure or viscous).
Upon integration by parts, it is obvious that the impulse encompasses the momentum of the fluid within the volume $\mathcal{V}$.
However, one can also show that the surface terms correspond to the momentum outside the volume $\mathcal{V}$ \citep[e.g.][]{Akhmetov2009}.
A thermal with volume $\mathcal{V}$, density $\rho$, and translating with velocity $\bm{u} = w_{\text{th}} \hat{z}$ thus has an impulse
\begin{equation}\label{eqn:momentum}
I_z = (1+k) \rho \mathcal{V} w_{\text{th}},
\end{equation}
where $k$ encompasses the ``virtual mass effect'' from the environmental fluid moving together with the thermal \citep{tarshish&all2018}.

We now restrict our study to an ideal gas in the low Mach-number regime, in an adiabatic background.
This is the appropriate regime of deep solar convection.
Due to rapid pressure equilibration in low Mach-number flows, we can approximate 
\begin{equation*}
\frac{\rho_1}{\rho_0} \approx -\frac{S_1}{c_P},
\end{equation*}
where $S_1$ is the specific entropy perturbation and $c_P$ is the specific heat at constant pressure; thermodynamic variables are decomposed into background (subscript 0) and fluctuating (subscript 1) components.

The rate of change of the impulse is
\begin{equation*}
\frac{d\bm{I}}{d t} = \int_\mathcal{V} \rho_1\, \bm{g}\, dV,
\end{equation*}
because the surface terms completely cancel.
Assuming a uniform, vertical gravity, $\bm{g} = -g\hat{z}$, it is useful to define the buoyancy perturbation,
\begin{equation}
B \equiv \int_{\mathcal{V}} \rho_0\, S_1\, \frac{g}{c_P}\, dV.
\label{eqn:tot_buoyancy}
\end{equation}
such that
\begin{equation}
\frac{d I_z}{d t} = B.
\label{eqn:change_in_impulse}
\end{equation}

@@

\subsection{Model of thermal evolution}
\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{constants.pdf}
    \caption{
	Shown is the evolution of buoyancy perturbation (top), circulation (middle), and volume factor (bottom) for each 2D simulation conducted in this work.
	All three of these quantities remain nearly constant after an initial spin-up phase.
	With increasing stratification, we see marginally more detrainment (loss of buoyant signature) in the top panel.
    \label{fig:constants} }
\end{figure}

The thermals studied here began as initial spherical perturbations and spin up into vortex rings.
The spun up vortex ring phase can be modeled as having evolved from a ``virtual origin'' where the vortex ring had zero radius.
We model the thermal as having been located at its virtual origin at a temporal offset $t = -t_{\text{off}}$, where $t = 0$ is the time at which the true thermal was released from rest.

In our simulations, we find the thermal undergoes weak detrainment (loss of buoyant signature to environmental fluid), so the negative buoyancy of the thermal decreases slightly in time (Fig.~\ref{fig:constants}, top panel).
We thus express the buoyancy perturbation as 
\begin{equation}
B \approx \eta B_{\text{th}},
\end{equation}
where $\eta$ is a constant of $\mathcal{O}$(1) which represents this detrainment and $B_{\text{th}}$ is the thermal's characteristic buoyancy perturbation. 
We then integrate Eqn.~\ref{eqn:change_in_impulse},
\begin{equation*}
I_z = \eta B_{\text{th}} (t + t_{\text{off}}).
\end{equation*}

Combining this with Eqn.~\ref{eqn:impulse_approx}, we retrieve our first main result,
\begin{equation}
r = \sqrt{\frac{\eta B_{\text{th}} (t + t_{\text{off}})}{\pi\rho_0\Gamma_{\text{th}}}},
\label{eqn:r_theory}
\end{equation}
Here, $\Gamma_{\text{th}}$ is the characteristic circulation of the thermal. As there are no net baroclinic torques, $\Gamma_{\text{th}}$ remains nearly constant in our simulations (Fig.~\ref{fig:constants}, middle panel).
In the Boussinesq limit where $\rho \rightarrow \text{constant}$, we retrieve the $r \propto \sqrt{t}$ scaling found in \LJ.
In the limit of strong stratification, we find $r \propto \rho^{-1/2}$, corresponding to purely horizontal compression, with $r^2 \propto \rho^{-1}$ \citep{brandenburg2016}.

To solve for the vertical evolution of the thermal, we can use Eqn.~\ref{eqn:momentum}.
We approximate the volume as $\mathcal{V} \approx m r^3$, where $m$ is a parameter which we take to be constant (which is a decent assumption in our simulations; see Fig.~\ref{fig:constants}, bottom panel).
Here, $m$ accounts for volumetric constants (e.g., $4\pi/3$), the aspect ratio of the thermal, and the cubed ratio between the full radius of the spheroidal thermal and $r$.
Defining the thermal velocity $w_{\text{th}} \equiv dz_{\text{th}}/dt$, we find
\begin{equation}
\frac{dz_{\text{th}}}{\rho(z_{\text{th}})^{1/2}} =
\left(\frac{(\pi\Gamma_{\text{th}})^{3/2}}{m(1 + k)(\eta B_{\text{th}})^{1/2}}\right)\frac{dt}{(t + t_{\text{off}})^{1/2}}.
\label{eqn:dz_theory}
\end{equation}
This can be easily integrated given an atmospheric stratification $\rho(z)$.

@@

\subsection{Polytropic atmosphere solution}
In this work, we study an ideal gas with an adiabatic index of $\gamma = 5/3$.
An adiabatic polytrope satisfies
\begin{gather}\label{eqn:T0}
T_0 = 1 + (z - L_z)\grad_{\text{ad}}, \\
\rho_0 = T_0^{\,n_{\text{ad}}},
\label{eqn:polytrope}
\end{gather}
where $n_{\text{ad}} = (\gamma-1)^{-1}$ and $\grad_{\text{ad}}$ is the adiabatic temperature gradient.
All thermodynamic quantities are nondimensionalized such that $\rho_0 = T_0 = 1$ at $z = L_z$, the top of the domain.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% EXPERIMENT SECTION
\section{Simulation setup} 
\label{sec:experiment}
To test our theory, we run a series of thermal simulations using the 3D fully compressible equations in cartesian domains.
We additionally compute 2D azimuthally-symmetric simulations using the anelastic equations in cylindrical domains.
We verify the 3D and 2D simulations produce the same results when run with the same parameters.
Because 2D simulations are less computationally expensive, we use them to cover a broader parameter regime.

@@

Following \LJ\,, we non-dimensionalize the equations on the initial diameter of the thermal and its freefall velocity.
These equations are then fully specified in terms of the Reynolds number and Prandtl number,
\begin{equation}
\text{Re} = \frac{ u_{\text{th}} L_{\text{th}}}{\nu}, \qquad
\text{Pr} = \frac{ u_{\text{th}} L_{\text{th}}}{\chi}, \qquad
u_{\text{th}}^2 = \frac{g L_{\text{th}} \Delta s}{c_P},
\end{equation}
where $u_{\text{th}}$ is the freefall velocity, $L_{\text{th}}$ is the thermal length scale, and
$\Delta s$ is the magnitude of the specific entropy signature of the thermal.

The background density and temperature are given by Eqs.~\ref{eqn:T0} \& \ref{eqn:polytrope}.
The adiabatic temperature gradient is $\nabla_{\text{ad}}=g(e^{N_\rho/n_{\text{ad}}}-1)/(L_z c_P)$, where $L_z=20$ is the height of the domain and $N_\rho$ is the number of density scale heights spanned by the domain.

We choose an atmospheric model in which the dynamic viscosity, $\mu = \rho_0 \nu$, and the thermal conductivity, $\kappa = \rho_0 \chi$, are both uniform in space and constant in time.
We make this choice because $\mu$ and $\kappa$ appear in the density-weighted momentum and entropy equations, and we find the density-weighted entropy and momentum to be key quantities in our thermal theory.
The diffusivities $\nu$ and $\chi$ therefore scale inversely with the density.
As the diffusivities scale with depth, Re is specified at the thermal's initial depth.
All simulations conducted in this work use an initial value of Re = 600 and $\mbox{\text{Pr} = 1}$.

\subsection{3D Fully Compressible Simulations}
In order to verify our 2D anelastic simulations, we also simulate thermals with the 3D Navier Stokes equations. 
We use the $(T, \ln\rho)$ formulation of the equations \citep{lecoanet&all2014, anders&brown2017},
\begin{gather}
\frac{\partial \ln\rho_1}{\partial t} + \epsilon^{-1}\left(\bm{u}\cdot\grad\ln\rho_0 + \DivU\right) = -\bm{u}\cdot\grad\ln\rho_1, \\
\begin{split}
\frac{\partial \bm{u}}{\partial t}  + \frac{1}{-\grad_\text{ad}}[\grad T_1 + &T_1\grad\ln\rho_0 + T_0\grad\ln\rho_1]  =\\
- \frac{\epsilon}{-\grad_\text{ad}} &T_1\grad\ln\rho_1 + \frac{1}{\rho\text{Re}}\left[\grad^2\bm{u} + \frac{1}{3}\grad(\DivU)\right]
\end{split} \\
\begin{split}
&\frac{\partial T_1}{\partial t} + \epsilon^{-1}\left[\bm{u}\cdot\grad T_0 + (\gamma-1)T_0\DivU\right] = \\
&-\left[\bm{u}\cdot\grad T_1 + (\gamma-1)T_1\DivU\right] + \frac{-\grad_{\text{ad}}}{\rho c_V\text{Re}}\left[\frac{1}{\text{Pr}}\grad^2 T_1 + \sigma_{ij}\partial_{x_i}u_j\right].
\end{split}
\end{gather}
These equations are non-dimensionalized in the same way as the anelastic equations, and use the same background atmosphere.
The new parameter $\epsilon = u_{\text{th}}^2$ is the magnitude of entropy perturbations and sets the Mach number of the thermal flows analagously to the adiabatic excess in polytropic convection \citep{anders&brown2017}; we use $\epsilon = 10^{-4}$ in this work.

@@

For the fully compressible simulations, we must also specify the density perturbation $\rho_1$.
We pick perturbations that are in pressure equilibrium,
\begin{equation}
\ln\rho_1 = S_1/c_P, \qquad T_1 = T_0(e^{-\epsilon\ln\rho_1} - 1)/\epsilon.
\end{equation}
In all cases, we do not add any symmetry breaking perturbations (e.g., noise).

@@

The 2D simulation domain is periodic in the z-direction with $z \in [-L_z/4, L_z]$ and the radial direction spans $r \in [0, L_r]$.
The boundary conditions are $\partial_r S_1 = w = (\grad\times\bm{u})_\phi = 0$ at $r = L_r$, and the regularity of the equations automatically impose $u = \partial_r(w) = \partial_r(S_1) = 0$ at $r = 0$.
The 3D simulation domain is periodic in the horizontal directions ($x, y \in [-L_r, L_r]$) and vertically spans $z \in [0, L_z]$.
We impose impenetrable, stress free, fixed-temperature boundary conditions at the upper and lower boundaries.
In all of our simulations we specify $L_z = 20$ and $L_r = 5$.
We extend our 2D simulation domains to $z = -L_z/4$ because those simulations are vertically periodic.
This extension allows us to study the full transit of the thermal above $z = 0$ and terminate the simulation before it begins to wrap through the bottom of the periodic domain.

In our 2D simulations, we represent the radial direction with Chebyshev polynomials so that we can include geometric factors in our equations and capture the singularity at $r = 0$.
We then assume the z direction is periodic to make the calculations more efficient, and we end the simulations before the thermal can interact with the vertical periodic boundaries.
Remarkably, the results of these simulations vary only minimally from our 3D simulations, as we will show in the next section, indicating that the vertical periodicity and the different side boundaries (cylinder vs. cube) do not influcence the thermal properties.

All of the code used to perform the simulations in this work can be found online in the supplementary materials in a Zenodo repository \citep{supp_andersetall2019} at \dataset[10.5281/zenodo.3311894]{https://doi.org/10.5281/zenodo.3311894}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% RESULTS SECTION
\section{Model verification}
\label{sec:results}
\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{results_panels.pdf}
    \caption{
	(top) The thermal depth as a function of time for the 2D anelastic (AN) and 3D fully compressible (FC) simulations.
	(bottom) The corresponding thermal velocities as a function of depth.
	Theoretical predictions from section \ref{sec:theory} are plotted in thin solid lines for each case (see parameters in table \ref{table:parameters}).
    \label{fig:results_panels} }
\end{figure}

\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{diff_AN_FC.pdf}
    \caption{
	(top) Thermal radius as a function of depth for the 2D anelastic (AN) and 3D fully compressible (FC) simulations.
	Theoretical predictions from section \ref{sec:theory} are plotted in thin solid lines.
	(bottom) The fractional difference between anelastic and fully compressible simulations.
    \label{fig:diff} }
\end{figure}
